---
title: "Exploratory analysis of TCGA COAD data"
author: "Kieran Campbell"
date: "9 October 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

library(scater)
library(tidyverse)
library(magrittr)
library(matrixStats)
library(ggrepel)
library(rstan)
library(MCMCglmm)
library(coda)
library(cowplot)
library(biomaRt)
library(Rcpp)
library(goseq)
library(viridis)

source("../gibbs/phenotime.R")
sourceCpp("../gibbs/phenotime.cpp")

set.seed(123L)

scale_vec <- function(x) (x - mean(x)) / sd(x)
```

# Data perparation

```{r data-prep, include = TRUE, message = FALSE}
load("../data/sce_coad_kallisto.Rdata")
sce <- sce[matrixStats::rowVars(tpm(sce)) > 0, ]

is_exprs(sce) <- tpm(sce) > 0


sce <- calculateQCMetrics(sce)


hgnc_symbols <- sapply(strsplit(featureNames(sce), "_"), `[`, 2)

dvar <- data_frame(var_exprs = matrixStats::rowVars(exprs(sce)),
                   hgnc_symbols, gene = featureNames(sce))
fData(sce) <- cbind(fData(sce), dvar)

set.seed(123L)
sce <- plotPCA(sce, ncomponents = 3,
               colour_by = "short_plate",
               return_SCESet = TRUE)


sc <- sce[, redDim(sce)[,1] < 0 & redDim(sce)[,3] > -10]

## After subsetting, re-calculate QC metrics
sc <- calculateQCMetrics(sc)
rm(sce) # can't store large sceset (??)
```

And let's plot this and colour by tumour / normal:

```{r sc-pca}
plotPCA(sc, colour_by = "sample_type", ncomponents = 3)
```

And create an SCESet using just the tumour samples:

```{r create tumour-only dataset}
sc_tumour <- sc[, sc$sample_type != "normal"]
```


# Selecting genes

We can select the high variance genes:

```{r select-highvar}
var_cutoff <- 7

filter(fData(sc_tumour), mean_exprs > 0.1, var_exprs > 0.1) %>% 
  ggplot(aes(x = mean_exprs, y = var_exprs, fill = n_cells_exprs)) +
  geom_point(shape = 21, alpha = 0.8) + 
  geom_hline(yintercept = var_cutoff, linetype = 2) +
  scale_fill_viridis(name = "Number\nSamples\nExpressed", option = "inferno") +
  theme(axis.title = element_text(size = 11)) + 
  ylab("Variance log expression") + xlab("Mean log expression")

```

What do these transcripts correspond to?

```{r transcripts-go-analysis}
# get ensembl gene ids
fData(sc_tumour)$ensembl_transcript_id <- 
  sapply(strsplit(featureNames(sc_tumour), ".", fixed = TRUE), `[`, 1)

fdata_for_go <- filter(fData(sc_tumour), var_exprs > 0.1, mean_exprs > 0.1)
highvar_transcripts <- fdata_for_go$ensembl_transcript_id[fdata_for_go$var_exprs > var_cutoff]


ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
ensembl_res <- getBM(c("ensembl_transcript_id", "ensembl_gene_id", "gene_biotype"), 
            filters = "ensembl_transcript_id",
            values = fdata_for_go$ensembl_transcript_id, mart = ensembl)

ensembl_res$is_highvar <- 1 * (ensembl_res$ensembl_transcript_id %in% highvar_transcripts)

gene_highvar <- group_by(ensembl_res, ensembl_gene_id) %>% 
  summarise(prop_highvar = mean(is_highvar)) %>% 
  mutate(is_gene_highvar = prop_highvar > 0)

gene_highvar <- inner_join(gene_highvar,
                           distinct(dplyr::select(ensembl_res, ensembl_gene_id, gene_biotype), ensembl_gene_id, gene_biotype),
                           by = "ensembl_gene_id")

filter(gene_highvar, is_gene_highvar) %>% 
  ggplot(aes(x = gene_biotype)) + geom_bar() +
  theme(axis.text.x = element_text(angle = 90))

go_input <- 1 * (gene_highvar$is_gene_highvar)
names(go_input) <- gene_highvar$ensembl_gene_id

pwf <- nullp(go_input, "hg38", "ensGene")
go <- goseq(pwf, "hg38", "ensGene")#, test.cats = "GO:BP")

go_tidy <- dplyr::select(go, over_represented_pvalue, term) %>% 
  mutate(qvalue = p.adjust(over_represented_pvalue, method = "BH"),
         log10_qvalue = -log10(qvalue)) %>% 
  head(n = 10) %>% 
  arrange((log10_qvalue)) %>% tbl_df()

go_tidy %>% 
  mutate(term = factor(as.character(term), levels = go_tidy$term)) %>% 
  ggplot(aes(x = term, y = log10_qvalue)) +
  geom_bar(stat = "identity") + coord_flip() +
  theme(axis.text.y = element_text(size = 7), axis.title.y = element_blank()) +
  xlab("-log10(q-val)") 
```


Now construct the covariates:

```{r construct-priors}
x_metastasis <- 1 * (sc_tumour$m_stage != "m0")
x_metastasis[is.na(x_metastasis)] <- mean(x_metastasis, na.rm = TRUE)
x_metastasis <- scale_vec(x_metastasis)

x_msi <- scale_vec( 1 * (sc_tumour$msi_status == "msi-h") )

x <- cbind(x_msi, x_metastasis)
```

We can see if they're correlated:

```{r are-covs-cor}
print( cor(x) )
```

See if there's anything on a pca plot:

```{r pca-plot}
sc_tumour$x_msi <- x_msi
sc_tumour$x_metastasis <- x_metastasis

# Plot PCA using only high-var transcripts
highvar_fnames <- featureNames(sc_tumour)[fData(sc_tumour)$var_exprs > var_cutoff]

plotPCA(sc_tumour, colour_by = "x_msi", ncomponents = 3, feature_set = highvar_fnames)
plotPCA(sc_tumour, colour_by = "x_metastasis", 
        ncomponents = 3, feature_set = highvar_fnames)
```

We see there are a couple of outliers on PC2 that we'll remove before continuing:

```{r remove-outliers-2}
set.seed(13)
sc_tumour <- plotPCA(sc_tumour, return_SCESet = TRUE, feature_set = highvar_fnames)
sc_tumour <- sc_tumour[, redDim(sc_tumour)[,2] < 12]
```

and plot the PCA again to see any difference:

```{r plot-pca}
 plotPCA(sc_tumour, colour_by = "x_msi", ncomponents = 3, feature_set = highvar_fnames)
plotPCA(sc_tumour, colour_by = "x_metastasis", 
        ncomponents = 3, feature_set = highvar_fnames)
```

And now we're good to fit phenotime:

```{r fit-phenotime}
x <- as.matrix(dplyr::select(pData(sc_tumour), x_msi, x_metastasis))
y <- scale(t(exprs(sc_tumour[highvar_fnames,])))
```

```{r gibbs-sample-lipid}
set.seed(123)
gibbs <- phenot(y, x[,,drop = F], iter = 2000, thin = 2)
```

Diagnostic plot:

```{r diagnostic-lipid}
qplot(seq_along(gibbs$lp_trace), gibbs$lp_trace, geom = 'line')
```

Then extract MAP ordering and compare to PCA:

```{r map-to-pca-nmd-lipid}
tmap <- colMeans(gibbs$pst_trace)
pc1 <- prcomp(y)$x[,1]
pc2 <- prcomp(y)$x[,2]

data_frame(tmap, pc1, msi = factor(x[,1])) %>% 
  ggplot(aes(x = pc1, y = tmap, color = msi)) + geom_point() +
  scale_color_brewer(palette = "Set1") + ylab("phenotime")

data_frame(tmap, pc1, msi = factor(x[,2])) %>% 
  ggplot(aes(x = pc1, y = tmap, color = msi)) + geom_point() +
  scale_color_brewer(palette = "Set1") + ylab("phenotime")
```


We also want to find genes that show an interaction:

```{r find-interaction-genes-lipid}
int_df <- data_frame(
  gene = colnames(y),
  tau_pg_1_mean = colMeans(gibbs$tau_pg_trace[,1,]),
  beta_1_mean = colMeans(gibbs$beta_trace[,1,]),
  tau_pg_2_mean = colMeans(gibbs$tau_pg_trace[,2,]),
  beta_2_mean = colMeans(gibbs$beta_trace[,2,]),
  c_mean = colMeans(gibbs$c_trace),
  alpha_mean = colMeans(gibbs$alpha_trace[,1,]),
  eta_map = colMeans(gibbs$eta_trace)
)

abs_beta_cutoff <- 0.15
tau_cutoff <- 59

ggplot(int_df, aes(x = abs(beta_1_mean), y = 1 / tau_pg_1_mean, fill = abs(c_mean))) +
  geom_point(shape = 21, size = 3, alpha = 0.9) + viridis::scale_fill_viridis(name = "MAP |c|") +
  xlab("Coefficient magnitude") + ylab("Posterior variance") +
  geom_hline(yintercept = 1 / tau_cutoff, linetype = 2, color = "grey") +
  geom_vline(xintercept = abs_beta_cutoff, linetype = 2, color = "grey")

ggplot(int_df, aes(x = abs(beta_2_mean), y = 1 / tau_pg_2_mean, fill = abs(c_mean))) +
  geom_point(shape = 21, size = 3, alpha = 0.9) + viridis::scale_fill_viridis(name = "MAP |c|") +
  xlab("Coefficient magnitude") + ylab("Posterior variance") +
  geom_hline(yintercept = 1 / tau_cutoff, linetype = 2, color = "grey") +
  geom_vline(xintercept = abs_beta_cutoff, linetype = 2, color = "grey")

ggplot(int_df, aes(x = c_mean, y = beta_1_mean, fill = tau_pg_1_mean)) +
  geom_point(shape = 21, size = 3, alpha = 0.9) + viridis::scale_fill_viridis(name = "MAP tau_pg") 

ggplot(int_df, aes(x = c_mean, y = beta_2_mean, fill = tau_pg_2_mean)) +
  geom_point(shape = 21, size = 3, alpha = 0.9) + viridis::scale_fill_viridis(name = "MAP tau_pg") 

ggplot(int_df, aes(x = beta_1_mean, y = beta_2_mean)) +
  geom_point(shape = 21, size = 3, alpha = 0.9)
```

We can plot gene expression along phenotime coloured by MSI status:

```{r expression_along_phenotime-lipid, fig.width = 10}
features_to_keep <- int_df$tau_pg_2_mean < tau_cutoff & abs(int_df$beta_2_mean) > abs_beta_cutoff

as_data_frame(y[, features_to_keep]) %>% 
  mutate(phenotime = tmap, msi = factor(x[,1])) %>% 
  gather(gene, expression, -phenotime, -msi) %>% 
  ggplot(aes(x = phenotime, y = expression, color = msi)) +
  geom_point(alpha = 0.6) + facet_wrap(~ gene, scales = "free_y") + 
  scale_color_brewer(name = "MSI-high", palette = "Set1")
```

And plot gene expression for common genes:

```{r common-genes}
features_to_keep <- abs(int_df$c_mean) > 1 & abs(int_df$beta_mean) < 0.01

as_data_frame(y[, features_to_keep]) %>% 
  mutate(phenotime = tmap, msi = factor(x[,1])) %>% 
  gather(gene, expression, -phenotime, -msi) %>% 
  ggplot(aes(x = phenotime, y = expression, color = msi)) +
  geom_point(alpha = 0.6) + facet_wrap(~ gene, scales = "free_y") + 
  scale_color_brewer(name = "MSI-high", palette = "Set1")
```

See if any concordance with stages:

```{r conc-with-stages}
data_frame(stage = factor(priors), phenotime = tmap) %>% 
  ggplot(aes(x = stage, y = phenotime)) + geom_violin() +
  geom_jitter(alpha = 0.5)
```

```{r days-to-death}
sc_tumour$phenotime <- tmap

pd_nonzero <- pData(sc_tumour) %>% 
  dplyr::select(phenotime, patient.days_to_death, msi_status) %>% 
  dplyr::filter(!is.na(patient.days_to_death), patient.days_to_death > 0)

ggplot(pd_nonzero, aes(x = phenotime, y = log10(patient.days_to_death),
                       color = msi_status)) +
  geom_point()


fit <- lm(log10(patient.days_to_death) ~ phenotime, data = pd_nonzero)
```

## Compare to MSI only

Then extract MAP ordering and compare to PCA:

```{r map-to-pca-nmd-lipid}
gibbs_msi_only <- phenot(y, x[,1,drop = F], iter = 2000, thin = 2)

tmap_msi <- colMeans(gibbs_msi_only$pst_trace)
pc1 <- prcomp(y)$x[,1]
pc2 <- prcomp(y)$x[,2]

data_frame(tmap_msi, pc1, msi = factor(x[,1])) %>% 
  ggplot(aes(x = pc1, y = tmap, color = msi)) + geom_point() +
  scale_color_brewer(palette = "Set1") + ylab("phenotime")

```

```{r msi-vs-2}
qplot(tmap_msi, tmap, color = factor(x[,1]))
```


We also want to find genes that show an interaction:

```{r find-interaction-genes-lipid}
int_df_msi <- data_frame(
  gene = colnames(y),
  tau_pg_1_mean = colMeans(gibbs_msi_only$tau_pg_trace[,1,]),
  beta_1_mean = colMeans(gibbs_msi_only$beta_trace[,1,]),
  c_mean = colMeans(gibbs_msi_only$c_trace),
  alpha_mean = colMeans(gibbs_msi_only$alpha_trace[,1,]),
  eta_map = colMeans(gibbs_msi_only$eta_trace)
)

ggplot(int_df_msi, aes(x = abs(beta_1_mean), y = 1 / tau_pg_1_mean, fill = abs(c_mean))) +
  geom_point(shape = 21, size = 3, alpha = 0.9) + viridis::scale_fill_viridis(name = "MAP |c|") +
  xlab("Coefficient magnitude") + ylab("Posterior variance") +
  geom_hline(yintercept = 1 / tau_cutoff, linetype = 2, color = "grey") +
  geom_vline(xintercept = abs_beta_cutoff, linetype = 2, color = "grey")


plot_grid(
  qplot(tmap, tmap_msi) +
    ggtitle("Pseudotime") + xlab("MSI + Meta") + ylab("MSI only"),
    qplot(int_df$c_mean, int_df_msi$c_mean) +
    ggtitle("Factor loading") + xlab("MSI + Meta") + ylab("MSI only"),
  qplot(int_df$beta_1_mean, int_df_msi$beta_1_mean) +
    ggtitle("Beta-MSI Coefficients") + xlab("MSI + Meta") + ylab("MSI only"),
  qplot(int_df$tau_pg_1_mean, int_df_msi$tau_pg_1_mean) +
    ggtitle("Beta-MSI Coefficient precision") + xlab("MSI + Meta") + ylab("MSI only")
)
```

# Metastasis-only

```{r metastasis-only}
set.seed(123)
gibbs_meta <- phenot(y, x[,2,drop = F], iter = 2000, thin = 2)
```

Diagnostic plot:

```{r diagnostic-lipid}
qplot(seq_along(gibbs_meta$lp_trace), gibbs$lp_trace, geom = 'line')
```

Then extract MAP ordering and compare to PCA:

```{r map-to-pca-nmd-lipid}
tmap_meta <- colMeans(gibbs_meta$pst_trace)
pc1 <- prcomp(y)$x[,1]
pc2 <- prcomp(y)$x[,2]

data_frame(tmap_meta, pc1, meta = factor(x[,2])) %>% 
  ggplot(aes(x = pc1, y = tmap, color = meta)) + geom_point() +
  scale_color_brewer(palette = "Set1") + ylab("phenotime")
```


We also want to find genes that show an interaction:

```{r find-interaction-genes-lipid}
int_df_meta <- data_frame(
  gene = colnames(y),
  tau_pg_2_mean = colMeans(gibbs_meta$tau_pg_trace[,1,]),
  beta_2_mean = colMeans(gibbs_meta$beta_trace[,1,]),
  c_mean = colMeans(gibbs_meta$c_trace),
  alpha_mean = colMeans(gibbs_meta$alpha_trace[,1,]),
  eta_map = colMeans(gibbs_meta$eta_trace)
)

abs_beta_cutoff <- 0.15
tau_cutoff <- 59


ggplot(int_df_meta, aes(x = abs(beta_2_mean), y = 1 / tau_pg_2_mean, fill = abs(c_mean))) +
  geom_point(shape = 21, size = 3, alpha = 0.9) + viridis::scale_fill_viridis(name = "MAP |c|") +
  xlab("Coefficient magnitude") + ylab("Posterior variance") +
  geom_hline(yintercept = 1 / tau_cutoff, linetype = 2, color = "grey") +
  geom_vline(xintercept = abs_beta_cutoff, linetype = 2, color = "grey")


ggplot(int_df, aes(x = c_mean, y = beta_2_mean, fill = tau_pg_2_mean)) +
  geom_point(shape = 21, size = 3, alpha = 0.9) + viridis::scale_fill_viridis(name = "MAP tau_pg") 
```

We can plot gene expression along phenotime coloured by MSI status:

```{r expression_along_phenotime-lipid, fig.width = 10}
features_to_keep <- int_df_meta$tau_pg_2_mean < tau_cutoff & abs(int_df_meta$beta_2_mean) > 0.25# abs_beta_cutoff

as_data_frame(y[, features_to_keep, drop = FALSE]) %>% 
  mutate(phenotime = tmap_meta, meta = factor(x[,2])) %>% 
  gather(gene, expression, -phenotime, -meta) %>% 
  filter(meta != 0) %>% 
  ggplot(aes(x = phenotime, y = expression, color = meta)) +
  geom_point(alpha = 0.5) + facet_wrap(~ gene, scales = "free_y") + 
  scale_color_brewer(name = "Metastasised", palette = "Set1") + 
  stat_smooth(se = FALSE, size = 1.8)


plot_grid(
  qplot(tmap, tmap_meta) +
    ggtitle("Pseudotime") + xlab("meta + Meta") + ylab("Meta only"),
    qplot(int_df$c_mean, int_df_meta$c_mean) +
    ggtitle("Factor loading") + xlab("meta + Meta") + ylab("Meta only"),
  qplot(int_df$beta_2_mean, int_df_meta$beta_2_mean) +
    ggtitle("Beta-meta Coefficients") + xlab("meta + Meta") + ylab("Meta only"),
  qplot(int_df$tau_pg_2_mean, int_df_meta$tau_pg_2_mean) +
    ggtitle("Beta-meta Coefficient precision") + xlab("meta + Meta") + ylab("Meta only")
)

tau_g_map <- colMeans(gibbs$tau_trace)
tau_g_map_meta <- colMeans(gibbs_meta$tau_trace)
tau_g_map_msi <- colMeans(gibbs_msi_only$tau_trace)

qplot(tau_g_map, tau_g_map_meta) + stat_function(fun = function(x) x) +
  xlab("tau_g MAP, x = [msi, meta]") +
  ylab("tau_g MAP, x = [meta]")

qplot(tau_g_map, tau_g_map_msi) + stat_function(fun = function(x) x)
```





