---
title: "Exploratory analysis of TCGA COAD data"
author: "Kieran Campbell"
date: "9 October 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

library(scater)
library(tidyverse)
library(magrittr)
library(matrixStats)
library(ggrepel)
library(rstan)
library(MCMCglmm)
library(coda)
library(cowplot)
library(biomaRt)
library(Rcpp)

source("../gibbs/phenotime.R")
sourceCpp("../gibbs/phenotime.cpp")

set.seed(123L)

scale_vec <- function(x) (x - mean(x)) / sd(x)
```

# Data perparation

```{r data-prep, include = TRUE, message = FALSE}
load("../data/sce_coad_kallisto.Rdata")
sce <- sce[matrixStats::rowVars(tpm(sce)) > 0, ]

is_exprs(sce) <- tpm(sce) > 0


sce <- calculateQCMetrics(sce)


hgnc_symbols <- sapply(strsplit(featureNames(sce), "_"), `[`, 2)

dvar <- data_frame(var_exprs = matrixStats::rowVars(exprs(sce)),
                   hgnc_symbols, gene = featureNames(sce))
fData(sce) <- cbind(fData(sce), dvar)

set.seed(123L)
sce <- plotPCA(sce, ncomponents = 3,
               colour_by = "short_plate",
               return_SCESet = TRUE)


sc <- sce[, redDim(sce)[,1] < 0 & redDim(sce)[,3] > -10]

## After subsetting, re-calculate QC metrics
sc <- calculateQCMetrics(sc)
rm(sce) # can't store large sceset (??)
```

And let's plot this and colour by tumour / normal:

```{r sc-pca}
plotPCA(sc, colour_by = "sample_type", ncomponents = 3)
```

And create an SCESet using just the tumour samples:

```{r create tumour-only dataset}
sc_tumour <- sc[, sc$sample_type != "normal"]
```

Finally construct the x matrix for microsatellite instability:

```{r construct-x}
msi_cov <- 1 * (sc_tumour$msi_status == "msi-h") 
x <- (matrix(msi_cov, ncol = 1))
```


# Using high-variance genes

Data prep:

```{r gibbs}
to_use <- fData(sc_tumour)$var_exprs > 6 # 7!!!
print(table(to_use))
y <- scale(t(exprs(sc_tumour[to_use,])))
```

Gibbs sample:

```{r gibbs-sample}
gibbs <- phenot(y, x, iter = 2000)
```

Diagnostic plot:

```{r diagnostic}
save(gibbs, file = "../data/gibbs_highvar.Rdata")
qplot(seq_along(gibbs$lp_trace), gibbs$lp_trace, geom = 'line')
```

Then extract MAP ordering and compare to PCA:

```{r map-to-pca}
tmap <- colMeans(gibbs$pst_trace)
pc1 <- prcomp(y)$x[,1]
pc2 <- prcomp(y)$x[,2]

data_frame(tmap, pc1, msi = factor(x[,1])) %>% 
  ggplot(aes(x = pc1, y = tmap, color = msi)) + geom_point() +
  scale_color_brewer(palette = "Set1") + ylab("phenotime")
```

We also want to find genes that show an interaction:

```{r find-interaction-genes}
int_df <- data_frame(
  tau_pg_map = matrixStats::colMedians(gibbs$tau_pg_trace[,1,]),
  beta_mean = colMeans(gibbs$beta_trace[,1,]),
  c_mean = colMeans(gibbs$c_trace)
)

abs_beta_cutoff <- 0.1
tau_cutoff <- 58

ggplot(int_df, aes(x = abs(beta_mean), y = 1 / tau_pg_map, fill = abs(c_mean))) +
  geom_point(shape = 21, size = 3, alpha = 0.9) + viridis::scale_fill_viridis(name = "MAP |c|") +
  xlab("Coefficient magnitude") + ylab("Posterior variance") +
  geom_hline(yintercept = 1 / tau_cutoff, linetype = 2, color = "grey") +
  geom_vline(xintercept = abs_beta_cutoff, linetype = 2, color = "grey")
```

We can plot gene expression along phenotime coloured by MSI status:

```{r expression_along_phenotime, fig.width = 10}
features_to_keep <- int_df$tau_pg_map < tau_cutoff & abs(int_df$beta_mean) > abs_beta_cutoff

as_data_frame(y[, features_to_keep]) %>% 
  mutate(phenotime = tmap, msi = factor(x[,1])) %>% 
  gather(gene, expression, -phenotime, -msi) %>% 
  ggplot(aes(x = phenotime, y = expression, color = msi)) +
  geom_point(alpha = 0.6) + facet_wrap(~ gene, scales = "free_y") + 
  scale_color_brewer(name = "MSI-high", palette = "Set1")
```

Plot selected features:

```{r selected-features, fig.width = 10, fig.height = 8}
genes_to_plot <- colnames(y)[features_to_keep][c(1:4, 9, 11)]

df_to_plot <- as_data_frame(y[, features_to_keep]) %>% 
  mutate(phenotime = tmap, msi = factor(x[,1])) %>% 
  gather(gene, expression, -phenotime, -msi)  

make_indiv_plot <- function(gtp) {
  d <- filter(df_to_plot, gene == gtp)
  ggplot(d, aes(x = phenotime, y = expression)) + 
    geom_point(data = dplyr::select(d, -msi), color = "grey") +
    geom_point(aes(x = phenotime, y = expression, color = msi)) +
    facet_wrap(~ msi, nrow = 2) + scale_color_brewer(palette = "Set1") +
    theme(legend.position = "none") +
    stat_smooth(method = "lm", se = FALSE, color = 'black', linetype = 2) +
    xlab("Phenotime") + ylab("Expression")
}

plot_list <- lapply(genes_to_plot, make_indiv_plot)
plot_grid(plotlist = plot_list, labels = sapply(strsplit(genes_to_plot, "_", fixed = TRUE), `[`, 2))
```



# Using NMD genes

Data prep:

```{r nmd-data-prep}
to_use <- fData(sc_tumour)$var_exprs > 2 &
  fData(sc_tumour)$gene_type == "nonsense_mediated_decay"
print(table(to_use))
y <- scale(t(exprs(sc_tumour[to_use,])))
```

Gibbs sample:

```{r gibbs-sample-nmd}
gibbs <- phenot(y, x, iter = 2000)
```

Diagnostic plot:

```{r diagnostic-nmd}
qplot(seq_along(gibbs$lp_trace), gibbs$lp_trace, geom = 'line')
```

Then extract MAP ordering and compare to PCA:

```{r map-to-pca-nmd}
tmap <- colMeans(gibbs$pst_trace)
pc1 <- prcomp(y)$x[,1]
pc2 <- prcomp(y)$x[,2]

data_frame(tmap, pc1, msi = factor(x[,1])) %>% 
  ggplot(aes(x = pc1, y = tmap, color = msi)) + geom_point() +
  scale_color_brewer(palette = "Set1") + ylab("phenotime")
```

We also want to find genes that show an interaction:

```{r find-interaction-genes-nmd}
int_df <- data_frame(
  tau_pg_map = matrixStats::colMedians(gibbs$tau_pg_trace[,1,]),
  beta_mean = colMeans(gibbs$beta_trace[,1,]),
  c_mean = colMeans(gibbs$c_trace)
)

abs_beta_cutoff <- 0.14
tau_cutoff <- 56

ggplot(int_df, aes(x = abs(beta_mean), y = 1 / tau_pg_map, fill = abs(c_mean))) +
  geom_point(shape = 21, size = 3, alpha = 0.9) + viridis::scale_fill_viridis(name = "MAP |c|") +
  xlab("Coefficient magnitude") + ylab("Posterior variance") +
  geom_hline(yintercept = 1 / tau_cutoff, linetype = 2, color = "grey") +
  geom_vline(xintercept = abs_beta_cutoff, linetype = 2, color = "grey")
```

We can plot gene expression along phenotime coloured by MSI status:

```{r expression_along_phenotime-nmd, fig.width = 10}
features_to_keep <- int_df$tau_pg_map < tau_cutoff & abs(int_df$beta_mean) > abs_beta_cutoff

as_data_frame(y[, features_to_keep]) %>% 
  mutate(phenotime = tmap, msi = factor(x[,1])) %>% 
  gather(gene, expression, -phenotime, -msi) %>% 
  ggplot(aes(x = phenotime, y = expression, color = msi)) +
  geom_point(alpha = 0.6) + facet_wrap(~ gene, scales = "free_y") + 
  scale_color_brewer(name = "MSI-high", palette = "Set1")
```




# Using lipid metabolism genes

Data prep:

```{r lipid-data-prep}
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
go_lipid <- "GO:0006629" # lipid metabolic process

la <- getBM(c("ensembl_transcript_id", "ensembl_gene_id"), filters = "go_id",
                   values = go_lipid, mart = ensembl)

fData(sc_tumour)$ensembl_transcript_id <- 
  sapply(strsplit(featureNames(sc_tumour), ".", fixed = TRUE), `[`, 1)
fData(sc_tumour)$is_lipid_transcript <- 
  fData(sc_tumour)$ensembl_transcript_id %in% la$ensembl_transcript_id

sc_lipid <- sc_tumour[fData(sc_tumour)$is_lipid_transcript &
                        fData(sc_tumour)$var_exprs > 0.1, ]



y <- scale(t(exprs(sc_lipid)))
```

Gibbs sample:

```{r gibbs-sample-lipid}
gibbs <- phenot(y, x, iter = 2000)
```

Diagnostic plot:

```{r diagnostic-lipid}
qplot(seq_along(gibbs$lp_trace), gibbs$lp_trace, geom = 'line')
```

Then extract MAP ordering and compare to PCA:

```{r map-to-pca-nmd-lipid}
tmap <- colMeans(gibbs$pst_trace)
pc1 <- prcomp(y)$x[,1]
pc2 <- prcomp(y)$x[,2]

data_frame(tmap, pc1, msi = factor(x[,1])) %>% 
  ggplot(aes(x = pc1, y = tmap, color = msi)) + geom_point() +
  scale_color_brewer(palette = "Set1") + ylab("phenotime")
```

We also want to find genes that show an interaction:

```{r find-interaction-genes-lipid}
int_df <- data_frame(
  tau_pg_map = matrixStats::colMedians(gibbs$tau_pg_trace[,1,]),
  beta_mean = colMeans(gibbs$beta_trace[,1,]),
  c_mean = colMeans(gibbs$c_trace)
)

abs_beta_cutoff <- 0.1
tau_cutoff <- 60

ggplot(int_df, aes(x = abs(beta_mean), y = 1 / tau_pg_map, fill = abs(c_mean))) +
  geom_point(shape = 21, size = 3, alpha = 0.9) + viridis::scale_fill_viridis(name = "MAP |c|") +
  xlab("Coefficient magnitude") + ylab("Posterior variance") +
  geom_hline(yintercept = 1 / tau_cutoff, linetype = 2, color = "grey") +
  geom_vline(xintercept = abs_beta_cutoff, linetype = 2, color = "grey")
```

We can plot gene expression along phenotime coloured by MSI status:

```{r expression_along_phenotime-lipid, fig.width = 10}
features_to_keep <- int_df$tau_pg_map < tau_cutoff & abs(int_df$beta_mean) > abs_beta_cutoff

as_data_frame(y[, features_to_keep]) %>% 
  mutate(phenotime = tmap, msi = factor(x[,1])) %>% 
  gather(gene, expression, -phenotime, -msi) %>% 
  ggplot(aes(x = phenotime, y = expression, color = msi)) +
  geom_point(alpha = 0.6) + facet_wrap(~ gene, scales = "free_y") + 
  scale_color_brewer(name = "MSI-high", palette = "Set1")
```

# With priors on the capture times


```{r construct-priors}
stage <- sc_lipid$stage

# parse stages
final_letter <- sapply(stage, function(s) str_sub(s, -1, -1))
to_trim <- final_letter %in% letters[1:3]
stage[to_trim] <- sapply(stage[to_trim], 
                                 function(s) str_sub(s, 1, -2))
stage <- toupper(sapply(stage, function(x) gsub("stage ", "", x)))

priors <- as.numeric(as.roman(stage))
priors[is.na(priors)] <- mean(priors, na.rm = TRUE)
priors <- scale_vec(priors)

## Metastasis priors
priors <- 1 * (sc_tumour$m_stage == "m0")
priors[is.na(priors)] <- mean(priors, na.rm = TRUE)

to_use <- fData(sc_tumour)$var_exprs > 7
print(table(to_use))
y <- scale(t(exprs(sc_tumour[to_use,])))
```

```{r gibbs-sample-lipid}
gibbs <- phenot(y, x, q = priors, tau_q = 1, iter = 1000)
```

Diagnostic plot:

```{r diagnostic-lipid}
qplot(seq_along(gibbs$lp_trace), gibbs$lp_trace, geom = 'line')
```

Then extract MAP ordering and compare to PCA:

```{r map-to-pca-nmd-lipid}
tmap <- colMeans(gibbs$pst_trace)
pc1 <- prcomp(y)$x[,1]
pc2 <- prcomp(y)$x[,2]

data_frame(tmap, pc1, msi = factor(x[,1])) %>% 
  ggplot(aes(x = pc1, y = tmap, color = msi)) + geom_point() +
  scale_color_brewer(palette = "Set1") + ylab("phenotime")

data_frame(tmap, pc1, stage = factor(x[,2])) %>% 
  ggplot(aes(x = pc1, y = tmap, color = stage)) + geom_point() +
  scale_color_brewer(palette = "Set1") + ylab("phenotime")
```


We also want to find genes that show an interaction:

```{r find-interaction-genes-lipid}
int_df <- data_frame(
  tau_pg_1_mean = matrixStats::colMedians(gibbs$tau_pg_trace[,1,]),
  beta_1_mean = colMeans(gibbs$beta_trace[,1,]),
  tau_pg_2_mean = matrixStats::colMedians(gibbs$tau_pg_trace[,2,]),
  beta_2_mean = colMeans(gibbs$beta_trace[,2,])
)

abs_beta_cutoff <- 0.15
tau_cutoff <- 59

ggplot(int_df, aes(x = abs(beta_1_mean), y = 1 / tau_pg_1_mean)) +
  geom_point(size = 3, alpha = 0.9) + 
  xlab("Coefficient magnitude") + ylab("Posterior variance") +
  geom_hline(yintercept = 1 / tau_cutoff, linetype = 2, color = "grey") +
  geom_vline(xintercept = abs_beta_cutoff, linetype = 2, color = "grey")


ggplot(int_df, aes(x = abs(beta_2_mean), y = 1 / tau_pg_2_mean)) +
  geom_point(size = 3, alpha = 0.9) + 
  xlab("Coefficient magnitude") + ylab("Posterior variance") +
  geom_hline(yintercept = 1 / tau_cutoff, linetype = 2, color = "grey") +
  geom_vline(xintercept = abs_beta_cutoff, linetype = 2, color = "grey")


```

We can plot gene expression along phenotime coloured by MSI status:

```{r expression_along_phenotime-lipid, fig.width = 10}
features_to_keep <- int_df$tau_pg_2_mean < tau_cutoff & abs(int_df$beta_2_mean) > abs_beta_cutoff

as_data_frame(y[, features_to_keep]) %>% 
  mutate(phenotime = tmap, msi = factor(x[,2])) %>% 
  gather(gene, expression, -phenotime, -msi) %>% 
  ggplot(aes(x = phenotime, y = expression, color = msi)) +
  geom_point(alpha = 0.6) + facet_wrap(~ gene, scales = "free_y") + 
  scale_color_brewer(name = "MSI-high", palette = "Set1")
```

And plot gene expression for common genes:

```{r common-genes}
features_to_keep <- abs(int_df$c_mean) > 1 & abs(int_df$beta_mean) < 0.01

as_data_frame(y[, features_to_keep]) %>% 
  mutate(phenotime = tmap, msi = factor(x[,1])) %>% 
  gather(gene, expression, -phenotime, -msi) %>% 
  ggplot(aes(x = phenotime, y = expression, color = msi)) +
  geom_point(alpha = 0.6) + facet_wrap(~ gene, scales = "free_y") + 
  scale_color_brewer(name = "MSI-high", palette = "Set1")
```

See if any concordance with stages:

```{r conc-with-stages}
data_frame(stage = factor(priors), phenotime = tmap) %>% 
  ggplot(aes(x = stage, y = phenotime)) + geom_violin() +
  geom_jitter(alpha = 0.5)
```

```{r days-to-death}
sc_tumour$phenotime <- tmap

pd_nonzero <- pData(sc_tumour) %>% 
  dplyr::select(phenotime, patient.days_to_death, msi_status) %>% 
  dplyr::filter(!is.na(patient.days_to_death), patient.days_to_death > 0)

ggplot(pd_nonzero, aes(x = phenotime, y = log10(patient.days_to_death),
                       color = msi_status)) +
  geom_point()


fit <- lm(log10(patient.days_to_death) ~ phenotime, data = pd_nonzero)
```


# With stage and msi as covariates


```{r construct-priors}
stage <- sc_lipid$stage

# parse stages
final_letter <- sapply(stage, function(s) str_sub(s, -1, -1))
to_trim <- final_letter %in% letters[1:3]
stage[to_trim] <- sapply(stage[to_trim], 
                                 function(s) str_sub(s, 1, -2))
stage <- toupper(sapply(stage, function(x) gsub("stage ", "", x)))

# x_stage <- as.numeric(as.roman(stage))
# x_stage[is.na(x_stage)] <- mean(x_stage, na.rm = TRUE)
# x_stage <- scale_vec(x_stage)

x_metastasis <- 1 * (sc_tumour$m_stage != "m0")
x_metastasis[is.na(x_metastasis)] <- mean(x_metastasis, na.rm = TRUE)
x_metastasis <- scale_vec(x_metastasis)

x_msi <- scale_vec( 1 * (sc_tumour$msi_status == "msi-h") )

x <- cbind(x_msi, x_metastasis)


to_use <- fData(sc_tumour)$var_exprs > 7
print(table(to_use))
y <- scale(t(exprs(sc_tumour[to_use,])))
```

```{r gibbs-sample-lipid}
set.seed(123)
gibbs <- phenot(y, x[,,drop = F], iter = 8000, thin = 4)
```

Diagnostic plot:

```{r diagnostic-lipid}
qplot(seq_along(gibbs$lp_trace), gibbs$lp_trace, geom = 'line')
```

Then extract MAP ordering and compare to PCA:

```{r map-to-pca-nmd-lipid}
tmap <- colMeans(gibbs$pst_trace)
pc1 <- prcomp(y)$x[,1]
pc2 <- prcomp(y)$x[,2]

data_frame(tmap, pc1, msi = factor(x[,1])) %>% 
  ggplot(aes(x = pc1, y = tmap, color = msi)) + geom_point() +
  scale_color_brewer(palette = "Set1") + ylab("phenotime")

data_frame(tmap, pc1, msi = factor(x[,2])) %>% 
  ggplot(aes(x = pc1, y = tmap, color = msi)) + geom_point() +
  scale_color_brewer(palette = "Set1") + ylab("phenotime")
```


We also want to find genes that show an interaction:

```{r find-interaction-genes-lipid}
int_df <- data_frame(
  gene = colnames(y),
  tau_pg_1_mean = matrixStats::colMedians(gibbs$tau_pg_trace[,1,]),
  beta_1_mean = colMeans(gibbs$beta_trace[,1,]),
  tau_pg_2_mean = matrixStats::colMedians(gibbs$tau_pg_trace[,2,]),
  beta_2_mean = colMeans(gibbs$beta_trace[,2,]),
  c_mean = colMeans(gibbs$c_trace),
  alpha_mean = colMeans(gibbs$alpha_trace[,1,]),
  eta_map = colMeans(gibbs$eta_trace)
)

abs_beta_cutoff <- 0.15
tau_cutoff <- 59

ggplot(int_df, aes(x = abs(beta_1_mean), y = 1 / tau_pg_1_mean, fill = abs(c_mean))) +
  geom_point(shape = 21, size = 3, alpha = 0.9) + viridis::scale_fill_viridis(name = "MAP |c|") +
  xlab("Coefficient magnitude") + ylab("Posterior variance") +
  geom_hline(yintercept = 1 / tau_cutoff, linetype = 2, color = "grey") +
  geom_vline(xintercept = abs_beta_cutoff, linetype = 2, color = "grey")

ggplot(int_df, aes(x = abs(beta_2_mean), y = 1 / tau_pg_2_mean, fill = abs(c_mean))) +
  geom_point(shape = 21, size = 3, alpha = 0.9) + viridis::scale_fill_viridis(name = "MAP |c|") +
  xlab("Coefficient magnitude") + ylab("Posterior variance") +
  geom_hline(yintercept = 1 / tau_cutoff, linetype = 2, color = "grey") +
  geom_vline(xintercept = abs_beta_cutoff, linetype = 2, color = "grey")

ggplot(int_df, aes(x = c_mean, y = beta_1_mean, fill = tau_pg_1_mean)) +
  geom_point(shape = 21, size = 3, alpha = 0.9) + viridis::scale_fill_viridis(name = "MAP tau_pg") 

ggplot(int_df, aes(x = c_mean, y = beta_2_mean, fill = tau_pg_2_mean)) +
  geom_point(shape = 21, size = 3, alpha = 0.9) + viridis::scale_fill_viridis(name = "MAP tau_pg") 

ggplot(int_df, aes(x = beta_1_mean, y = beta_2_mean)) +
  geom_point(shape = 21, size = 3, alpha = 0.9)
```

We can plot gene expression along phenotime coloured by MSI status:

```{r expression_along_phenotime-lipid, fig.width = 10}
features_to_keep <- int_df$tau_pg_2_mean < tau_cutoff & abs(int_df$beta_2_mean) > abs_beta_cutoff

as_data_frame(y[, features_to_keep]) %>% 
  mutate(phenotime = tmap, msi = factor(x[,1])) %>% 
  gather(gene, expression, -phenotime, -msi) %>% 
  ggplot(aes(x = phenotime, y = expression, color = msi)) +
  geom_point(alpha = 0.6) + facet_wrap(~ gene, scales = "free_y") + 
  scale_color_brewer(name = "MSI-high", palette = "Set1")
```

And plot gene expression for common genes:

```{r common-genes}
features_to_keep <- abs(int_df$c_mean) > 1 & abs(int_df$beta_mean) < 0.01

as_data_frame(y[, features_to_keep]) %>% 
  mutate(phenotime = tmap, msi = factor(x[,1])) %>% 
  gather(gene, expression, -phenotime, -msi) %>% 
  ggplot(aes(x = phenotime, y = expression, color = msi)) +
  geom_point(alpha = 0.6) + facet_wrap(~ gene, scales = "free_y") + 
  scale_color_brewer(name = "MSI-high", palette = "Set1")
```

See if any concordance with stages:

```{r conc-with-stages}
data_frame(stage = factor(priors), phenotime = tmap) %>% 
  ggplot(aes(x = stage, y = phenotime)) + geom_violin() +
  geom_jitter(alpha = 0.5)
```

```{r days-to-death}
sc_tumour$phenotime <- tmap

pd_nonzero <- pData(sc_tumour) %>% 
  dplyr::select(phenotime, patient.days_to_death, msi_status) %>% 
  dplyr::filter(!is.na(patient.days_to_death), patient.days_to_death > 0)

ggplot(pd_nonzero, aes(x = phenotime, y = log10(patient.days_to_death),
                       color = msi_status)) +
  geom_point()


fit <- lm(log10(patient.days_to_death) ~ phenotime, data = pd_nonzero)
```



