---
title: "Exploratory analysis of TCGA OV data"
author: "Kieran Campbell"
date: "9 October 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

library(scater)
library(tidyverse)
library(magrittr)
library(matrixStats)
library(ggrepel)
library(rstan)
library(MCMCglmm)
library(coda)

scale_vec <- function(x) (x - mean(x)) / sd(x)
```


```{r data-prep, include = FALSE}
load("../data/sce_coad_kallisto.Rdata")
sce <- sce[matrixStats::rowVars(tpm(sce)) > 0, ]
is_exprs(sce) <- tpm(sce) > 0
sce <- calculateQCMetrics(sce)

hgnc_symbols <- sapply(strsplit(featureNames(sce), "_"), `[`, 2)

dvar <- data_frame(var_exprs = matrixStats::rowVars(exprs(sce)),
                   hgnc_symbols, gene = featureNames(sce))
fData(sce) <- cbind(fData(sce), dvar)

## group factors to avoid low frequency ones
gt_table <- table(fData(sce)$gene_type)
new_gt_names <- names(gt_table)
new_gt_names[gt_table < 50] <- "other"
fData(sce)$short_gene_type <- plyr::mapvalues(fData(sce)$gene_type,
                                             from = names(gt_table),
                                             to = new_gt_names)
```

# Strip out technical effects

```{r pca-all}
sce <- plotPCA(sce, ncomponents = 3, colour_by = "short_plate",
               return_SCESet = TRUE)
```

We see technically distinct clusters of cells. We want to target the largest, so strip out those that aren't there:

```{r subset-sceset}
sc <- sce[, redDim(sce)[,1] < 0 & redDim(sce)[,3] > -10]

## After subsetting, re-calculate QC metrics
sc <- calculateQCMetrics(sc)
```

And let's plot this and colour by tumour / normal:

```{r sc-pca}
plotPCA(sc, colour_by = "sample_type", ncomponents = 3)
```

# Mean-variance relationship

Mean variance relationships are usually informative. 

```{r mean-var}
filter(fData(sc), var_exprs > 1) %>% 
  ggplot(aes(x = mean_exprs, y = var_exprs, color = short_gene_type)) + 
    geom_point() + xlab("Mean") +
    ylab("Variance") +
    geom_text_repel(data = filter(fData(sc), var_exprs > 15), 
                    aes(label = gene_type), size = 5)

filter(fData(sc), var_exprs > 1) %>% 
  ggplot( aes(x = mean_exprs, y = var_exprs)) + 
    geom_point(alpha = 0.5) + xlab("Mean") +
    ylab("Variance") +
    geom_point(data = filter(fData(sc), var_exprs > 1,
                             gene_type == "nonsense_mediated_decay"), 
                    aes(label = hgnc_symbols, fill = short_gene_type), 
               color = 'black', shape = 21, size = 3)

filter(fData(sc), var_exprs > 1) %>% 
  ggplot(aes(x = mean_exprs, y = var_exprs)) + 
  geom_point(alpha = 0.5) + xlab("Mean") +
  ylab("Variance") +
  geom_point(data = filter(fData(sc), gene_type == "IG_V_gene", var_exprs > 1), 
             aes(label = hgnc_symbols, fill = short_gene_type), 
             color = 'black', shape = 21, size = 3)
```


# Technical effects of ACTB

We have concerns about the technical distribution of one ACTB transcript:

```{r actb-plot}
plotExpression(sc, feature = "ENST00000464611.1_ACTB", x = "plate")
```

Let's see what the extent of correlations with ACTB is and how that compares to a null distribution:

```{r actb-cors}
actb_exprs <- exprs(sc)["ENST00000464611.1_ACTB", ]
cors_with_actb <- as.numeric(cor(t(exprs(sc)), actb_exprs))
null_cors_with_actb <- as.numeric(cor(t(exprs(sc)), sample(actb_exprs)))

data_frame(actb = cors_with_actb, null = null_cors_with_actb) %>% 
  gather(type, correlation) %>% 
  ggplot(aes(x = correlation, fill = type)) + geom_density(alpha = 0.7) 
```




# What are the PCs?

Finally, we assign each cell a IG-V score and a nonsense-mediated decay score (by simply taking the mean expression of the genes associated with these for each cell).

```{r assign-scores}
get_mean_exprs <- function(term) {
  pathway_genes <- filter(fData(sc), gene_type %in% term) %>% extract2("gene")
  pathway_expression <- t(exprs(sc)[pathway_genes, ])
  pathway_expression <- pathway_expression[,colVars(pathway_expression) > 0]
  mean_exprs <- rowMeans(scale(pathway_expression))
  return(mean_exprs)
}

sc$nmd_score <- get_mean_exprs("nonsense_mediated_decay")
sc$igv_score <- get_mean_exprs("IG_V_gene")

ggplot(pData(sc), aes(x = censored, y = igv_score)) + geom_boxplot()
ggplot(pData(sc), aes(x = censored, y = nmd_score)) + geom_boxplot()
```

First, are the scores correlated?

```{r are-correlated}
plotPhenoData(sc, aes_string(x = "nmd_score", y = "igv_score"))
```

Nope! What happens if we plot PCA by these scores:

```{r pca-by-score, fig.width = 10, fig.height = 9}
plotPCA(sc, colour_by = "pct_dropout", ncomponents = 4)
plotPCA(sc, colour_by = "igv_score", ncomponents = 4)
plotPCA(sc, colour_by = "nmd_score", ncomponents = 4)
```

So PC1 = dropout, PC2 = IG-V, PC4 = NMD. Cool!

What about the NMD pathway being associated with somatic mutations

```{r nmd-somatic}
plotPhenoData(sc, aes_string(x = "n_mutations", y = "nmd_score"))
```

# TREGs and CTLs

```{r tregs}
ctl4 <- "ENSG00000163599"

treg_marker_ind <- featureNames(sc)[match("CD4", fData(sc)$hgnc_symbol)]

sc$cd4_expression <- exprs(sc)[treg_marker_ind,]


fit_igv <- lm(sc$cd4_expression ~ sc$igv_score)
print(paste("R2:", summary(fit_igv)$r.squared))

ggplot(pData(sc), aes_string(x = "igv_score", y = "cd4_expression")) +
  geom_point() + ggtitle(expression(paste(R^2, " = 0.57"))) +
  ylab("CD4 expression (TREGs)")

fit_nmd <- lm(sc$cd4_expression ~ sc$nmd_score)
print(paste("R2:", summary(fit_nmd)$r.squared))

ggplot(pData(sc), aes_string(x = "nmd_score", y = "cd4_expression")) +
  geom_point() + ggtitle(expression(paste(R^2, " = 0.09"))) +
  ylab("CD4 expression (TREGs)")



## There's a foxp3 transcript that shows correlated expression
sc$foxp3 <- exprs(sc)["ENST00000376207.8_FOXP3", ]

plotPhenoData(sc, aes_string(x = "cd4_expression", y = "foxp3"))
plotPhenoData(sc, aes_string(x = "igv_score", y = "foxp3"))

sc$treg_score <- rowMeans(cbind(scale_vec(sc$foxp3), scale_vec(sc$cd4_expression)))

plotPCA(sc, colour_by = "foxp3", ncomponents = 3)
```


# Time for some phenotime

Data preparation:

```{r phenotime-data-preparation}
sc_tumour <- sc[, sc$sample_type != "normal"]


weight_covariate <- as.numeric(sc_tumour$weight)
weight_covariate[is.na(weight_covariate)] <- mean(weight_covariate, na.rm = TRUE)

m_cov <- 1 * (sc_tumour$m_stage == "m0")
m_cov[is.na(m_cov)] <- round(mean(m_cov, na.rm = TRUE))

n_cov <- 1 * (sc_tumour$n_stage == "n0")
n_cov[is.na(m_cov)] <- round(mean(n_cov, na.rm = TRUE))

X <- rbind(m_cov, n_cov)#,
           # scale_vec(weight_covariate))
           
# X[is.na(X)] <- which.max(table(X)) - 1 # set 4 missing vals to mode

# let's focus on igv genes
genes_to_use <- filter(fData(sc_tumour),# gene_type %in% "nonsense_mediated_decay",
                    var_exprs > 10) %>% extract2("gene")
Y <- exprs(sc_tumour)[genes_to_use, ]
Y <- t( scale(t(Y)) )

N <- ncol(Y)
stopifnot(N == ncol(X))

G <- nrow(Y); P = nrow(X)

dlist <- list(N = N, G = G, P = P, X = X, Y = Y)
```



```{r model-fitting}
model <- stan_model("../stan/phenot_ard.stan",
                    model_name = "phenotime_ard")

fit <- vb(model, dlist, grad_samples = 3)
fit <- sampling(model, dlist, iter = 6000, chains = 1, thin = 2)
```

Look at results - 

```{r stan-results}
phenotime <- posterior.mode(mcmc(extract(fit, "pst")$pst))

pt_df <- data_frame(phenotime, 
                    igv_pca = prcomp(t(Y))$x[,1],
                     igv_score = sc_tumour$igv_score, foxp3 = sc_tumour$foxp3,
                     metastasis = factor(X[1, ]),
                      lymph = factor(X[2, ]))#, weight = X[3,])

ggplot(pt_df, aes(x = phenotime, y = igv_pca, colour = metastasis)) + geom_point() 
ggplot(pt_df, aes(x = phenotime, y = igv_pca, colour = lymph)) + geom_point() 


ggplot(pt_df, aes(x = phenotime, y = igv_pca, color = weight)) +
  geom_point() + viridis::scale_color_viridis() 

ggplot(pt_df, aes(x = phenotime, y = foxp3, color = metastasis)) +
  geom_point()

ggplot(pt_df, aes(x = phenotime, y = foxp3, color = weight)) +
  geom_point() + viridis::scale_color_viridis() 

```

Let's check some posteriors:

```{r check-posterior}
tidy_beta <- function(beta, n) {
  names(beta) <- rownames(Y)
  beta_tidy <- gather(beta, gene, value)
  beta_tidy$beta <- n
  return(beta_tidy)
}

beta_1_tidy <- tidy_beta( as_data_frame( extract(fit, "beta")$beta[,,1] ), "beta_1" )

ggplot(beta_1_tidy, aes(x = gene, y = value)) + geom_boxplot()

## Look at factor loadings

c_trace <- as_data_frame(extract(fit, "c")$c)
names(c_trace) <- rownames(Y)
c_tidy <- gather(c_trace, gene, value)

ggplot(c_tidy, aes(x = gene, y = value)) + geom_boxplot()

# and ARD posteriors for beta1
ard_covs <- apply(extract(fit, "cov_tau")$cov_tau, 3, 
                  function(x) posterior.mode(mcmc(x))) %>% 
  as_data_frame()
#names(ard_covs) <- "Metastasis" # rownames(X)

ard_cov_tidy <- mutate(ard_covs, gene = rownames(Y)) %>% 
  gather(covariate, value, -gene)

ggplot(ard_cov_tidy, aes(x = gene, y = value)) +
  geom_bar(stat = 'identity') + facet_wrap(~ covariate, ncol = 1)
```

Compare beta to c:

```{r beta-to-c}
bc <- data_frame(
  c = posterior.mode(mcmc(extract(fit, "c")$c)),
  beta = posterior.mode(mcmc(extract(fit, "beta")$beta[,,2]))
)

ggplot(bc, aes(x = c, y = beta)) + geom_point()
```


```{r check-genes}
m <- factor(X[1,])
n <- factor(X[2,])
weight <- weight_covariate


genes_reg <- filter(ard_cov_tidy, covariate == "V3",
                    value < 35) %>% extract2("gene") 

df_gex <- as_data_frame(t(Y[genes_reg, , drop = FALSE])) %>% 
  mutate(phenotime, weight) %>% 
  gather(gene, expression, -phenotime, -weight)

ggplot(df_gex, aes(x = phenotime, y = expression, color = weight )) +
  geom_point() + facet_wrap(~ gene, scales = "free_y") +
  scale_color_viridis()


genes_reg <- filter(ard_cov_tidy, covariate == "V1",
                    value < 20) %>% extract2("gene") 

df_gex <- as_data_frame(t(Y[genes_reg, , drop = FALSE])) %>% 
  mutate(phenotime, m) %>% 
  gather(gene, expression, -phenotime, -m)

ggplot(df_gex, aes(x = phenotime, y = expression, color = m )) +
  geom_point() + facet_wrap(~ gene, scales = "free_y")


genes_reg <- filter(ard_cov_tidy, covariate == "V2",
                    value < 20) %>% extract2("gene") 

df_gex <- as_data_frame(t(Y[genes_reg, , drop = FALSE])) %>% 
  mutate(phenotime, n) %>% 
  gather(gene, expression, -phenotime, -n)

ggplot(df_gex, aes(x = phenotime, y = expression, color = n )) +
  geom_point() + facet_wrap(~ gene, scales = "free_y")

```

Compare beta to c:

```{r beta-to-c}
ggplot(df_ard, aes(x = c_map, y = beta_map, color = is_igv_gene)) + geom_point()
```

# Cheating

```{r cheat-time}
fd <- filter(fData(sc_tumour), var_exprs > 2.5)
m_status <- sc_tumour$m

pvals <- apply(exprs(sc_tumour)[fd$gene,], 1, function(y) {
  fit <- lm(y ~ m_status)
  pval <- coef(summary(fit))[2,4]
  return(pval)
})

fd$m_pval <- pvals

pval_per_type <- group_by(fd, short_gene_type) %>% 
  summarise(mean_pval = median(m_pval), num = n())

```











# Learning TCGA with mutation data


## Explore mutation data


```{r explore-mutation-data}
ov_mutations <- filter(OV.mutations, bcr_patient_barcode != "bcr_patient_barcode")
num_represented_patients <- length(unique(ov_mutations$bcr_patient_barcode))

num_mutations <- group_by(ov_mutations, gene) %>% 
  summarise(prop_patients = n() / num_represented_patients)

num_mutations %>% filter(prop_patients > 0.05) %>% 
  arrange(prop_patients) %>% 
  ggplot(aes(x = gene, y = prop_patients)) + geom_bar(stat = "identity") +
  xlab("Mutated gene") + ylab("Proportion of patients with mutation") +
  ggtitle("Proportion of patients with mutation (>5% freq)")

mutations_per_gene <- group_by(ov_mutations, gene, bcr_patient_barcode) %>% 
  summarise(mpg = n())

filter(mutations_per_gene, mpg > 1) %>% 
  group_by(gene) %>% summarise(patients_with_gt_1_mutation = n()) %>% View()
```

## Model fitting


Data preparation:

```{r data-prep}
sc_somatic <- sc[, !is.na(sc$n_somatic)]

X <- model.matrix(~ n_mutations, pData(sc_somatic))
X <- t( scale( X[,-1] ) ) # remove intercept

# let's focus on igv genes
# igv_genes <- filter(fData(sc_somatic), gene_type %in% "IG_V_gene") %>% extract2("gene")
# Y <- exprs(sc_somatic)[igv_genes, ]
# Y <- t( scale(t(Y)) )

to_use <- matrixStats::rowVars(exprs(sc_somatic)) > 8
Y <- t( scale(t(exprs(sc_somatic)[to_use, ])))
sc_used <- sc_somatic[to_use, ]

# let's look at NMD genes
sc_nmd <- sc_somatic[fData(sc_somatic)$gene_type == "nonsense_mediated_decay", ]
Y <- exprs(sc_nmd)[rowVars(exprs(sc_nmd)) > 2.5, ]
Y <- t(scale(t(Y)))

N <- ncol(Y)
stopifnot(N == ncol(X))

G <- nrow(Y); P = nrow(X)

data_somatic <- list(N = N, G = G, P = P, X = X, Y = Y)

```


Fit model:

```{r model-fitting-somatic}
model <- stan_model("../stan/phenot.stan",
                    model_name = "phenotime")
model_ard <- stan_model("../stan/phenot_ard.stan",
                       model_name = "phenotime_ard")

fit <- vb(model_ard, data_somatic, grad_samples = 3)

fit <- sampling(model_ard, data_somatic, chains = 1, iter = 4000, thin = 4)
```

Extract results:

```{r extract-somatic-results}
phenotime <- posterior.mode(mcmc(extract(fit, "pst")$pst))

pt_df <- data_frame(phenotime,# pc1 = redDim(sc_somatic)[,1], 
                    #pc2 = redDim(sc_somatic)[,2], 
                    purepc1 = prcomp(t(Y), scale = TRUE)$x[,1],
                    purepc2 = prcomp(t(Y), scale = TRUE)$x[,2],
                     #igv_score = sc_somatic$igv_score, foxp3 = sc_somatic$foxp3,
                    n_mutations = sc_somatic$n_mutations)
```

Some plots:

```{r somatic-plots}
ggplot(pt_df, aes(x = scale_vec(purepc1), y = scale_vec(phenotime), 
                  color = n_mutations)) + geom_point() +
  viridis::scale_color_viridis() + stat_function(fun = function(x) x, linetype = 2)

ggplot(pt_df, aes(x = phenotime, y = foxp3, color = n_mutations)) + geom_point() +
  viridis::scale_color_viridis() 

ggplot(pt_df, aes(x = purepc1, y = foxp3, color = n_mutations)) + geom_point() +
  viridis::scale_color_viridis() 

```

Check plate effect:

```{r plate-effect}
data_frame(phenotime, plate = sc_nmd$plate) %>% ggplot(aes(x = plate, y = phenotime, fill = plate)) + geom_violin() + geom_jitter()
```

Check stage effect:

```{r stage-effect}
data_frame(phenotime, stage = sc_somatic$patient.stage_event.clinical_stage) %>% 
  ggplot(aes(x = stage, y = phenotime, color = stage)) + geom_violin() +
  geom_jitter() + theme(legend.position = 'none')
```

And look at the betas:

```{r check-betas}
beta_1_tidy <- tidy_beta( as_data_frame( extract(fit, "beta")$beta[,,1] ), "beta_1" ) %>% 
  mutate(is_igv_gene = gene %in% igv_genes)

ggplot(beta_1_tidy, aes(x = gene, y = value, fill = is_igv_gene)) + geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# and ARD posteriors for beta1
ard_cov <- posterior.mode(mcmc(extract(fit, "cov_tau")$cov_tau[,,1]))
df_ard <- data_frame(ard_cov, gene = rownames(Y))

df_ard$c_map <- posterior.mode(mcmc(extract(fit, "c")$c))
df_ard$beta_map <- posterior.mode(mcmc(extract(fit, "beta")$beta[,,1]))
df_ard %<>% mutate(is_igv_gene = gene %in% igv_genes)


ggplot(df_ard, aes(x = gene, y = ard_cov, fill = is_igv_gene)) +
  geom_bar(stat = 'identity') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

stan_hist(fit, c("a_beta", "b_beta"))

## look at the alphas
alpha_1_tidy <- tidy_beta( as_data_frame( extract(fit, "alpha")$alpha[,,1] ), "alpha_1" ) %>% 
  mutate(is_igv_gene = gene %in% igv_genes)

ggplot(alpha_1_tidy, aes(x = gene, y = value, fill = is_igv_gene)) + geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

Check which genes appear to be "regulated" by somatic mutations:

```{r check-genes}
genes_reg <- filter(df_ard) %>% extract2("gene") # , ard_cov < 10)

nmut <- sc_somatic$n_mutations

df_gex <- as_data_frame(t(Y)) %>% 
  mutate(phenotime, n_mutations = nmut) %>% 
  gather(gene, expression, -phenotime, -n_mutations)

ggplot(df_gex, aes(x = phenotime, y = expression, color = n_mutations)) +
  geom_point() + facet_wrap(~ gene, scales = "free_y") + 
  viridis::scale_color_viridis()

# genes_no_reg <- filter(df_ard, ard_cov > 600, abs(c_map) > 0.04) %>% extract2("gene")
# 
# df_gex_noreg <- as_data_frame(t(Y[genes_no_reg,])) %>% 
#   mutate(phenotime, n_somatic = sc_somatic$n_somatic) %>% 
#   gather(gene, expression, -phenotime, -n_somatic)
# 
# ggplot(df_gex_noreg, aes(x = phenotime, y = expression, color = n_somatic)) +
#   geom_point() + facet_wrap(~ gene, scales = "free_y") + 
#   viridis::scale_color_viridis()
```

Compare beta to c:

```{r beta-to-c}
ggplot(df_ard, aes(x = c_map, y = beta_map, color = is_igv_gene)) + geom_point()
```

Mean mutations along pseudotime:

```{r rolling-mean}
mutations_by_phenotime <- arrange(pt_df, phenotime) %>% extract2("n_mutations")
rolling_median <- zoo::rollmedian(mutations_by_phenotime, 101)
plot(rolling_median)

cowplot::plot_grid(ggplot(pt_df, aes(x = phenotime, y = n_mutations)) + geom_point(),
          ggplot(pt_df, aes(x = purepc1, y = n_mutations)) + geom_point())

lfit <- lm(n_mutations ~ poly(phenotime, 2), data = pt_df)
```


# Mutational signature as a phenotype


```{r mut-sig}
ov_mutations <- filter(OV.mutations, bcr_patient_barcode != "bcr_patient_barcode") %>% as_tibble()

ovm_tidy <- select(ov_mutations, gene, bcr_patient_barcode) %>% 
  mutate(has_mutation = 1) %>% 
  distinct() %>% 
  filter(gene != "TP53", gene != "TTN")

n_mutations <- group_by(ovm_tidy, gene) %>% summarise(n = n())
genes_to_keep <- filter(n_mutations, n > 5) %>% extract2("gene")

ovm_spread <- filter(ovm_tidy, gene %in% genes_to_keep) %>% 
  spread(gene, has_mutation, fill = 0)

mutation_matrix <- as.matrix(ovm_spread[,-1])
colnames(mutation_matrix) <- names(ovm_spread)[-1]
rownames(mutation_matrix) <- ovm_spread$bcr_patient_barcode

mutation_pca <- prcomp(mutation_matrix)

# let's check the gene loadings for the first PC
data_frame(gene = colnames(mutation_matrix), loading = mutation_pca$rotation[,1]) %>% 
  ggplot(aes(x = gene, y = loading)) + geom_bar(stat = 'identity')
```

# Appendix

## Prior on $\beta$

If we select tau_beta from gamma(10, 0.01) then we get a prior on beta like

```{r prior-on-beta}
qplot(rnorm(10000, 0, 1 / sqrt(rgamma(10000, 1000, 1))), geom = "density")
```

which seems reasonable.