---
title: "Exploratory analysis of TCGA OV data"
author: "Kieran Campbell"
date: "9 October 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

library(scater)
library(tidyverse)
library(magrittr)
library(matrixStats)
library(ggrepel)
library(rstan)
library(MCMCglmm)
library(coda)

scale_vec <- function(x) (x - mean(x)) / sd(x)
```


```{r data-prep, include = FALSE}
load("../data/sce_ovarian_kallisto.Rdata")
sc <- sce[matrixStats::rowVars(exprs(sce)) > 1, ]
is_exprs(sc) <- exprs(sc) > 0
sc <- calculateQCMetrics(sc)

hgnc_symbols <- sapply(strsplit(featureNames(sc), "_"), `[`, 2)

dvar <- data_frame(# mean_exprs = rowMeans(exprs(sc)),
                   var_exprs = matrixStats::rowVars(exprs(sc)),
                   hgnc_symbols, gene = featureNames(sc))
fData(sc) <- cbind(fData(sc), dvar)

## group factors to avoid low frequency ones
gt_table <- table(fData(sc)$gene_type)
new_gt_names <- names(gt_table)
new_gt_names[gt_table < 50] <- "other"
fData(sc)$short_gene_type <- plyr::mapvalues(fData(sc)$gene_type,
                                             from = names(gt_table),
                                             to = new_gt_names)
```

# Mean-variance relationship

Mean variance relationships are usually informative. Below are three plots of mean-variance, first coloured by gene biotype, then emphasising those involved in nonsense mediated decay and immunoglobuin-V:

```{r mean-var}
ggplot(fData(sc), aes(x = mean_exprs, y = var_exprs, color = short_gene_type)) + 
  geom_point() + xlab("Mean") +
  ylab("Variance") +
  geom_text_repel(data = filter(fData(sc), var_exprs > 8.5), 
                  aes(label = hgnc_symbols), size = 5)

ggplot(fData(sc), aes(x = mean_exprs, y = var_exprs)) + 
  geom_point(alpha = 0.5) + xlab("Mean") +
  ylab("Variance") +
  geom_point(data = filter(fData(sc), gene_type == "nonsense_mediated_decay"), 
                  aes(label = hgnc_symbols, fill = short_gene_type), 
             color = 'black', shape = 21, size = 3)

ggplot(fData(sc), aes(x = mean_exprs, y = var_exprs)) + 
  geom_point(alpha = 0.5) + xlab("Mean") +
  ylab("Variance") +
  geom_point(data = filter(fData(sc), gene_type == "IG_V_gene"), 
             aes(label = hgnc_symbols, fill = short_gene_type), 
             color = 'black', shape = 21, size = 3)
```


# Technical effects of ACTB

We have concerns about the technical distribution of one ACTB transcript:

```{r actb-plot}
plotExpression(sc, feature = "ENST00000464611.1_ACTB", x = "plate")
```

Let's see what the extent of correlations with ACTB is and how that compares to a null distribution:

```{r actb-cors}
actb_exprs <- exprs(sc)["ENST00000464611.1_ACTB", ]
cors_with_actb <- as.numeric(cor(t(exprs(sc)), actb_exprs))
null_cors_with_actb <- as.numeric(cor(t(exprs(sc)), sample(actb_exprs)))

data_frame(actb = cors_with_actb, null = null_cors_with_actb) %>% 
  gather(type, correlation) %>% 
  ggplot(aes(x = correlation, fill = type)) + geom_density(alpha = 0.7) 
```

We see there are a significant number of genes with large correlations (and anti-correlations) to ACTB. Let's have a look at the highest and lowest correlations:

```{r pickout-actb-genes}
plotExpression(sc, x = "plate", feature = featureNames(sc)[order(cors_with_actb, decreasing = T)][2])
plotExpression(sc, x = "plate", feature = featureNames(sc)[order(cors_with_actb)][1])
```

There appears to be technical effects here too. We can look at the correlations of genes with ACTB compared to there variance, and see how this varies with gene type:

```{r cor-var-plot, fig.width = 12, fig.height = 7}
fData(sc)$correlation_with_actb <- as.numeric(cors_with_actb)

ggplot(fData(sc), aes(x = correlation_with_actb, y = var_exprs,
                      color = short_gene_type)) +
  geom_point() + stat_density_2d(data  = filter(fData(sc), short_gene_type %in% c("nonsense_mediated_decay", "IG_V_gene")))
```

We see there are a number of high-variance genes highly correlated with ACTB, but the IG-V and NMD pathways largely avoid this (slight anti-correlation?).



# What are the PCs?

Finally, we assign each cell a IG-V score and a nonsense-mediated decay score (by simply taking the mean expression of the genes associated with these for each cell).

```{r assign-scores}
get_mean_exprs <- function(term) {
  pathway_genes <- filter(fData(sc), gene_type %in% term) %>% extract2("gene")
  mean_exprs <- rowMeans(scale(t(exprs(sc)[pathway_genes, ])))
  return(mean_exprs)
}

sc$nmd_score <- get_mean_exprs("nonsense_mediated_decay")
sc$igv_score <- get_mean_exprs("IG_V_gene")

ggplot(pData(sc), aes(x = censored, y = igv_score)) + geom_boxplot()
ggplot(pData(sc), aes(x = censored, y = nmd_score)) + geom_boxplot()
```

First, are the scores correlated?

```{r are-correlated}
plotPhenoData(sc, aes_string(x = "nmd_score", y = "igv_score"))
```

Nope! What happens if we plot PCA by these scores:

```{r pca-by-score, fig.width = 10, fig.height = 9}
plotPCA(sc, colour_by = "pct_dropout", ncomponents = 4)
plotPCA(sc, colour_by = "igv_score", ncomponents = 4)
plotPCA(sc, colour_by = "nmd_score", ncomponents = 4)
```

So PC1 = dropout, PC2 = IG-V, PC4 = NMD. Cool!

What about the NMD pathway being associated with somatic mutations

```{r nmd-somatic}
plotPhenoData(sc, aes_string(x = "n_mutations", y = "nmd_score"))
```

# TREGs and CTLs

```{r tregs}
ctl4 <- "ENSG00000163599"

treg_marker_ind <- featureNames(sc)[match("CD4", fData(sc)$hgnc_symbol)]

sc$cd4_expression <- exprs(sc)[treg_marker_ind,]

fit_igv <- lm(sc$cd4_expression ~ sc$igv_score)
print(paste("R2:", summary(fit_igv)$r.squared))

ggplot(pData(sc), aes_string(x = "igv_score", y = "cd4_expression")) +
  geom_point() + ggtitle(expression(paste(R^2, " = 0.24"))) +
  ylab("CD4 expression (TREGs)")

fit_nmd <- lm(sc$cd4_expression ~ sc$nmd_score)
print(paste("R2:", summary(fit_nmd)$r.squared))

ggplot(pData(sc), aes_string(x = "nmd_score", y = "cd4_expression")) +
  geom_point() + ggtitle(expression(paste(R^2, " = 0.16"))) +
  ylab("CD4 expression (TREGs)")


plotPCA(sc, colour_by = treg_marker_ind)
plotTSNE(sc, colour_by = treg_marker_ind, perplexity = 4)



## There's a foxp3 transcript that shows correlated expression
sc$foxp3 <- exprs(sce)["ENST00000376207.8_FOXP3", ]

plotPhenoData(sc, aes_string(x = "cd4_expression", y = "foxp3"))


sc$treg_score <- rowMeans(cbind(scale_vec(sc$foxp3), scale_vec(sc$cd4_expression)))

plotPCA(sc, colour_by = "foxp3", ncomponents = 3)
```


# Time for some phenotime

Data preparation:

```{r phenotime-data-preparation}

sc$patient.age_at_initial_pathologic_diagnosis <-
  as.numeric(sc$patient.age_at_initial_pathologic_diagnosis)

# sc$random <- sample(c(0,1), size = ncol(sc), replace = TRUE)
sc$random <- rnorm(ncol(sc))

X <- model.matrix(~ patient.age_at_initial_pathologic_diagnosis, pData(sc))
X <- model.matrix(~ random, pData(sc))
X <- t( scale( X[,-1] ) ) # remove intercept

# let's focus on igv genes
igv_genes <- filter(fData(sc), gene_type %in% "IG_V_gene") %>% extract2("gene")
Y <- exprs(sc)[igv_genes, ]
Y <- t( scale(t(Y)) )

N <- ncol(Y)
stopifnot(N == ncol(X))

G <- nrow(Y); P = nrow(X)

dlist <- list(N = N, G = G, P = P, X = X, Y = Y)
```



```{r model-fitting}
model <- stan_model("../stan/phenot.stan",
                    model_name = "phenotime")

fit <- vb(model, dlist, grad_samples = 3)
fit <- sampling(model, dlist, chains = 1, iter = 4000)
```

Look at results - 

```{r stan-results}
phenotime <- posterior.mode(mcmc(extract(fit, "pst")$pst))

pt_df <- data_frame(phenotime, pc1 = redDim(sc)[,1], pc2 = redDim(sc)[,2], 
                    igv_pca = prcomp(t(Y))$x[,1],
                     age = sc$patient.age_at_initial_pathologic_diagnosis,
                     igv_score = sc$igv_score, foxp3 = sc$foxp3,
                     censored = X[1, ], random = as.numeric(sc$random))

ggplot(pt_df, aes(x = phenotime, y = igv_pca, color = random)) + geom_point() +
  viridis::scale_color_viridis()

ggplot(pt_df, aes(x = phenotime, y = igv_pca, color = age)) +
  geom_point() + viridis::scale_color_viridis() 

ggplot(pt_df, aes(x = phenotime, y = foxp3, color = age)) +
  geom_point() + viridis::scale_color_viridis() 

ggplot(pt_df, aes(x = phenotime, y = pc1, color = age)) +
  geom_point() + viridis::scale_color_viridis() 

```

Let's check some posteriors:

```{r check-posterior}
tidy_beta <- function(beta, n) {
  names(beta) <- rownames(Y)
  beta_tidy <- gather(beta, gene, value)
  beta_tidy$beta <- n
  return(beta_tidy)
}

beta_1_tidy <- tidy_beta( as_data_frame( extract(fit, "beta")$beta[,,1] ), "beta_1" )

ggplot(beta_1_tidy, aes(x = gene, y = value)) + geom_boxplot()



## Look at factor loadings

c_trace <- as_data_frame(extract(fit, "c")$c)
names(c_trace) <- rownames(Y)
c_tidy <- gather(c_trace, gene, value)

ggplot(c_tidy, aes(x = gene, y = value)) + geom_boxplot()

# and ARD posteriors for beta1
ard_cov <- posterior.mode(mcmc(extract(fit, "cov_tau")$cov_tau[,,1]))

ggplot(data_frame(ard_cov, gene = rownames(Y)), aes(x = gene, y = ard_cov)) +
  geom_bar(stat = 'identity')
```


# Learning TCGA with mutation data

Data preparation:

```{r data-prep}
sc_somatic <- sc[, !is.na(sc$n_somatic)]

plotPhenoData(sc_somatic, aes_string(x = "patient.age_at_initial_pathologic_diagnosis",
                              y = "n_mutations"))

X <- model.matrix(~ n_mutations, pData(sc_somatic))
X <- t( scale( X[,-1] ) ) # remove intercept

# let's focus on igv genes
# igv_genes <- filter(fData(sc_somatic), gene_type %in% "IG_V_gene") %>% extract2("gene")
# Y <- exprs(sc_somatic)[igv_genes, ]
# Y <- t( scale(t(Y)) )

to_use <- matrixStats::rowVars(exprs(sc_somatic)) > 7
Y <- t( scale(t(exprs(sc_somatic)[to_use, ])))

# let's look at NMD genes
# sc_nmd <- sc_somatic[fData(sc_somatic)$gene_type == "nonsense_mediated_decay", ]
# Y <- exprs(sc_nmd)[rowVars(exprs(sc_nmd)) > 2.5, ]
# Y <- t(scale(t(Y)))

N <- ncol(Y)
stopifnot(N == ncol(X))

G <- nrow(Y); P = nrow(X)

data_somatic <- list(N = N, G = G, P = P, X = X, Y = Y)

```


Fit model:

```{r model-fitting-somatic}
model <- stan_model("../stan/phenot.stan",
                    model_name = "phenotime")
model_ard <- stan_model("../stan/phenot_ard.stan",
                       model_name = "phenotime_ard")

fit <- vb(model_ard, data_somatic, grad_samples = 3)

fit <- sampling(model_ard, data_somatic, chains = 1, iter = 2000, thin = 2)
```

Extract results:

```{r extract-somatic-results}
phenotime <- posterior.mode(mcmc(extract(fit, "pst")$pst))

pt_df <- data_frame(phenotime, pc1 = redDim(sc_nmd)[,1], 
                    pc2 = redDim(sc_nmd)[,2], 
                    purepc1 = prcomp(t(Y))$x[,1],
                     igv_score = sc_nmd$igv_score, foxp3 = sc_nmd$foxp3,
                    n_mutations = sc_nmd$n_mutations)
```

Some plots:

```{r somatic-plots}
ggplot(pt_df, aes(x = phenotime, y = purepc1, color = n_mutations)) + geom_point() +
  viridis::scale_color_viridis() + ylab("PC1")

ggplot(pt_df, aes(x = phenotime, y = foxp3, color = n_mutations)) + geom_point() +
  viridis::scale_color_viridis() 

```

And look at the betas:

```{r check-betas}
beta_1_tidy <- tidy_beta( as_data_frame( extract(fit, "beta")$beta[,,1] ), "beta_1" ) %>% 
  mutate(is_igv_gene = gene %in% igv_genes)

ggplot(beta_1_tidy, aes(x = gene, y = value, fill = is_igv_gene)) + geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# and ARD posteriors for beta1
ard_cov <- posterior.mode(mcmc(extract(fit, "cov_tau")$cov_tau[,,1]))
df_ard <- data_frame(ard_cov, gene = rownames(Y))

df_ard$c_map <- posterior.mode(mcmc(extract(fit, "c")$c))
df_ard$beta_map <- posterior.mode(mcmc(extract(fit, "beta")$beta[,,1]))
df_ard %<>% mutate(is_igv_gene = gene %in% igv_genes)


ggplot(df_ard, aes(x = gene, y = ard_cov, fill = is_igv_gene)) +
  geom_bar(stat = 'identity') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

stan_hist(fit, c("a_beta", "b_beta"))
```

Check which genes appear to be "regulated" by somatic mutations:

```{r check-genes}
genes_reg <- filter(df_ard) %>% extract2("gene") # , ard_cov < 10)

df_gex <- as_data_frame(t(Y)) %>% 
  mutate(phenotime, n_mutations = sc_somatic$n_mutations) %>% 
  gather(gene, expression, -phenotime, -n_mutations)

ggplot(df_gex, aes(x = phenotime, y = expression, color = n_mutations)) +
  geom_point() + facet_wrap(~ gene, scales = "free_y") + 
  viridis::scale_color_viridis()

# genes_no_reg <- filter(df_ard, ard_cov > 600, abs(c_map) > 0.04) %>% extract2("gene")
# 
# df_gex_noreg <- as_data_frame(t(Y[genes_no_reg,])) %>% 
#   mutate(phenotime, n_somatic = sc_somatic$n_somatic) %>% 
#   gather(gene, expression, -phenotime, -n_somatic)
# 
# ggplot(df_gex_noreg, aes(x = phenotime, y = expression, color = n_somatic)) +
#   geom_point() + facet_wrap(~ gene, scales = "free_y") + 
#   viridis::scale_color_viridis()
```

Compare beta to c:

```{r beta-to-c}
ggplot(df_ard, aes(x = c_map, y = beta_map, color = is_igv_gene)) + geom_point()
```

Mean mutations along pseudotime:

```{r rolling-mean}
mutations_by_phenotime <- arrange(pt_df, phenotime) %>% extract2("n_mutations")
rolling_median <- zoo::rollmedian(mutations_by_phenotime, 101)
plot(rolling_median)

ggplot(pt_df, aes(x = phenotime, y = n_mutations)) + geom_point() +
  stat_smooth(method = "lm")

lfit <- lm(n_mutations ~ phenotime, data = pt_df)
```


# Appendix

## Prior on $\beta$

If we select tau_beta from gamma(10, 0.01) then we get a prior on beta like

```{r prior-on-beta}
qplot(rnorm(10000, 0, 1 / sqrt(rgamma(10000, 1000, 1))), geom = "density")
```

which seems reasonable.