---
title: "Exploratory analysis of TCGA OV data"
author: "Kieran Campbell"
date: "9 October 2016"
output: html_document
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

library(scater)
library(tidyverse)
library(magrittr)
library(matrixStats)
library(ggrepel)
library(rstan)
library(MCMCglmm)
library(coda)
library(biomaRt)
library(GO.db)
library(viridis)
library(gplots)
library(RColorBrewer)
library(sva)


# Some functions to use
scale_vec <- function(x) (x - mean(x)) / sd(x)

tidy_beta <- function(beta, n) {
  names(beta) <- rownames(Y)
  beta_tidy <- gather(beta, gene, value)
  beta_tidy$beta <- n
  return(beta_tidy)
}
```


```{r data-prep, include = FALSE}
load("../data/sce_ovarian_kallisto.Rdata")
sc <- sce[matrixStats::rowVars(exprs(sce)) > 0, ]
is_exprs(sc) <- exprs(sc) > 0
sc <- calculateQCMetrics(sc)

hgnc_symbols <- sapply(strsplit(featureNames(sc), "_"), `[`, 2)

dvar <- data_frame(# mean_exprs = rowMeans(exprs(sc)),
                   var_exprs = matrixStats::rowVars(exprs(sc)),
                   hgnc_symbols, gene = featureNames(sc))
fData(sc) <- cbind(fData(sc), dvar)

## group factors to avoid low frequency ones
gt_table <- table(fData(sc)$gene_type)
new_gt_names <- names(gt_table)
new_gt_names[gt_table < 50] <- "other"
fData(sc)$short_gene_type <- plyr::mapvalues(fData(sc)$gene_type,
                                             from = names(gt_table),
                                             to = new_gt_names)

# Need to add ensembl transcript ids to SCESet:

ensembl_transcript_ids <- sapply(strsplit(featureNames(sc), ".", fixed = TRUE), `[`, 1)
fData(sc)$ensembl_transcript_id <- ensembl_transcript_ids

sc$foxp3 <- exprs(sce)["ENST00000376207.8_FOXP3", ]
```

# Gene annotations

Load genes associated with "lipid metabolism":

```{r load-lipid-met}
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
go_lipid <- "GO:0006629" # lipid metabolic process

xx <- as.list(GOBPCHILDREN)
all_lipid_children <- xx[[go_lipid]]

lipid_annotation <- getBM("ensembl_transcript_id", filters = "go_id",
                   values = all_lipid_children, mart = ensembl)

fData(sc)$is_lipid_transcript <- 
  fData(sc)$ensembl_transcript_id %in% lipid_annotation$ensembl_transcript_id
```

Let's remove the batch effect:

```{r remove-batch}
sc$plate <- droplevels(sc$plate)
batch <- sc$plate
design <- model.matrix(~ 1, data = pData(sc))

exprs_combat <- ComBat(exprs(sc), batch, design)
exprs(sc) <- exprs_combat
```


```{r sce-lippid}
sce_lipid <- sc[fData(sc)$is_lipid_transcript, ]
```

Subset to those with mutations:

```{r sce-lipid2}
sc_lipid <- sce_lipid[, !is.na(sce_lipid$n_mutations)]
sc_lipid <- sc_lipid[rowVars(exprs(sc_lipid)) > 0, ]
```

Quick heatmap of lipids:

```{r heatmap}
prop_cells_exprs <- rowMeans(exprs(sc_lipid) > 0)
sc_lipid <- sc_lipid[prop_cells_exprs > 0.2,]# & prop_cells_exprs < 1]
exprs_for_heatmap <- exprs(sc_lipid) 
# exprs_for_heatmap[exprs_for_heatmap > 5] <- 5
col_side_col <- plyr::mapvalues(sc_lipid$censored, from = c(TRUE, FALSE),
                                to = brewer.pal(3, "Set1")[1:2])
heatmap.2(exprs_for_heatmap, trace = "none", col = "viridis",
          ColSideColors = col_side_col)
```

Quick look at Foxp3

```{r foxp3-lipids}
lipid_pc1 <- prcomp(scale(t(exprs(sc_lipid))))$x[,1]

df <- data_frame(pc1 = lipid_pc1, foxp3 = sc_lipid$foxp3, 
                 pct_genes_expr = sc_lipid$pct_dropout) 

df %>% 
  ggplot(aes(x = -pc1, y = foxp3)) + geom_point() +
  xlab("PC1 from lipid metabolism") + ylab("Foxp3 expression")

df %>% 
  ggplot(aes(x = pct_genes_expr, y = foxp3)) + geom_point() +
  xlab("% genes expressed") + ylab("Foxp3 expression")

df %>% 
  ggplot(aes(x = pct_genes_expr, y = pc1)) + geom_point() +
  xlab("% genes expressed") + ylab("FPC1")

```

# Phenotime


```{r phenotime}
X <- model.matrix(~ foxp3, pData(sc_lipid))
X <- t( scale( X[,-1] ) ) # remove intercept

Y <- t(scale(t(exprs(sc_lipid))))

N <- ncol(Y)
stopifnot(N == ncol(X))

G <- nrow(Y); P = nrow(X)

data_lipid <- list(N = N, G = G, P = P, X = X, Y = Y)

```



Fit model:

```{r model-fitting-somatic}
model_ard <- stan_model("../stan/phenot_ard.stan",
                       model_name = "phenotime_ard")

fit <- vb(model_ard, data_lipid, grad_samples = 3, output_samples = 500)
```

Extract results:

```{r extract-somatic-results}
phenotime <- posterior.mode(mcmc(extract(fit, "pst")$pst))

pt_df <- data_frame(phenotime, foxp3 = sc_lipid$foxp3,
                    purepc1 = prcomp(t(Y), scale = TRUE)$x[,1],
                    purepc2 = prcomp(t(Y), scale = TRUE)$x[,2],
                    n_mutations = sc_lipid$n_mutations)
```

Some plots:

```{r somatic-plots}
ggplot(pt_df, aes(x = scale_vec(purepc1), y = -scale_vec(phenotime), 
                  color = foxp3)) + geom_point() +
  viridis::scale_color_viridis() + stat_function(fun = function(x) x, linetype = 2)

ggplot(pt_df, aes(x = phenotime, y = foxp3, color = n_mutations)) + geom_point() +
  viridis::scale_color_viridis() 

ggplot(pt_df, aes(x = purepc1, y = foxp3, color = n_mutations)) + geom_point() +
  viridis::scale_color_viridis() 

```

Check plate effect:

```{r plate-effect}
data_frame(phenotime, plate = sc_lipid$plate) %>% 
  ggplot(aes(x = plate, y = phenotime, fill = plate)) + 
  geom_violin() + geom_jitter()
```

Check stage effect:

```{r stage-effect}
data_frame(phenotime, stage = sc_lipid$patient.stage_event.clinical_stage) %>% 
  ggplot(aes(x = stage, y = phenotime, color = stage)) + geom_violin() +
  geom_jitter() + theme(legend.position = 'none')
```

And look at the betas:

```{r check-betas}
beta_1_tidy <- tidy_beta( as_data_frame( extract(fit, "beta")$beta[,,1] ), "beta_1" ) 

ggplot(beta_1_tidy, aes(x = gene, y = value)) + geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# and ARD posteriors for beta1
ard_cov <- posterior.mode(mcmc(extract(fit, "cov_tau")$cov_tau[,,1]))
df_ard <- data_frame(ard_cov, gene = rownames(Y))

df_ard$c_map <- posterior.mode(mcmc(extract(fit, "c")$c))
df_ard$beta_map <- posterior.mode(mcmc(extract(fit, "beta")$beta[,,1]))

ggplot(df_ard, aes(x = gene, y = ard_cov)) +
  geom_bar(stat = 'identity') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 4))

stan_hist(fit, c("a_beta", "b_beta"))

## look at the alphas
alpha_1_tidy <- tidy_beta( as_data_frame( extract(fit, "alpha")$alpha[,,1] ), "alpha_1" ) 

ggplot(alpha_1_tidy, aes(x = gene, y = value)) + geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

Check which genes appear to be "regulated" by somatic mutations:

```{r check-genes}
genes_reg <- arrange(df_ard, ard_cov) %>% extract2("gene") %>% head(n = 4)

nmut <- sc_lipid$n_mutations
foxp3 <- sc_lipid$foxp3

df_gex <- as_data_frame(t(Y[genes_reg, ,drop = FALSE])) %>% 
  mutate(phenotime, foxp3, nmut) %>% 
  gather(gene, expression, -phenotime, -foxp3, -nmut)

ggplot(df_gex, aes(x = phenotime, y = expression, color = foxp3)) +
  geom_point() + facet_wrap(~ gene, scales = "free_y") + 
  viridis::scale_color_viridis()

# genes_no_reg <- filter(df_ard, ard_cov > 600, abs(c_map) > 0.04) %>% extract2("gene")
# 
# df_gex_noreg <- as_data_frame(t(Y[genes_no_reg,])) %>% 
#   mutate(phenotime, n_somatic = sc_somatic$n_somatic) %>% 
#   gather(gene, expression, -phenotime, -n_somatic)
# 
# ggplot(df_gex_noreg, aes(x = phenotime, y = expression, color = n_somatic)) +
#   geom_point() + facet_wrap(~ gene, scales = "free_y") + 
#   viridis::scale_color_viridis()
```

Compare beta to c:

```{r beta-to-c}
ggplot(df_ard, aes(x = c_map, y = beta_map)) + geom_point()
```

Mean mutations along pseudotime:

```{r rolling-mean}
mutations_by_phenotime <- arrange(pt_df, phenotime) %>% extract2("n_mutations")
rolling_median <- zoo::rollmedian(mutations_by_phenotime, 101)
plot(rolling_median)

cowplot::plot_grid(ggplot(pt_df, aes(x = phenotime, y = n_mutations)) + geom_point(),
          ggplot(pt_df, aes(x = purepc1, y = n_mutations)) + geom_point())

lfit <- lm(n_mutations ~ poly(phenotime, 2), data = pt_df)
```


# Some quick cheating


```{r cheating}
pdata <- dplyr::select(pData(sc_lipid), patient.days_to_death,
                patient.age_at_initial_pathologic_diagnosis,
                # patient.stage_event.clinical_stage, censored,
                n_mutations, censored * 1)
pdata$patient.age_at_initial_pathologic_diagnosis <- 
  as.numeric(pdata$patient.age_at_initial_pathologic_diagnosis)

cors <- apply(pdata, 2, function(x) cor(x, phenotime, use = "complete.obs"))

```